---
title: "IMO Data Analysis and Aggregation"
subtitle: "Comprehensive Analysis of International Mathematical Olympiad Performance"
author: "Zhongyue Lin"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    code-fold: show
    theme: cosmo
---

## Introduction

This analysis examines IMO performance patterns across three dimensions:
- Individual country performance trends
- Award distribution patterns
- Participation and team composition analysis

## Setup and Data Import

```{r setup, message=FALSE, warning=FALSE}
# Load required libraries
library(tidyverse)  # Data manipulation
library(arrow)      # Parquet file handling
library(curl)       # Data download
library(ggrepel)    # Text label positioning
library(scales)     # Better plot scales
library(patchwork)  # Plot composition

# Configure proxy settings
Sys.setenv(http_proxy = "http://127.0.0.1:1087",
           https_proxy = "http://127.0.0.1:1087")

# Create output directory
dir.create("finalOutput", showWarnings = FALSE)

# Data import function with type handling
read_imo_data <- function(file_name, integer_cols, character_cols, date_cols = c()) {
  base_url <- "https://raw.githubusercontent.com/Zhongyuelin/Deliverable01-Collecting-raw-data-/main/IMO_data/"
  url <- paste0(base_url, file_name)
  
  h <- new_handle()
  handle_setopt(h, ssl_verifypeer = FALSE)
  temp <- tempfile()
  curl_download(url, temp, handle = h)
  
  col_types <- cols(.default = col_double())
  col_types$cols[integer_cols] <- map(integer_cols, ~ col_integer())
  col_types$cols[character_cols] <- map(character_cols, ~ col_character())
  col_types$cols[date_cols] <- map(date_cols, ~ col_date())
  
  data <- read_csv(temp, col_types = col_types)
  unlink(temp)
  return(data)
}

# Define color palette for consistent visualization
palette_main <- c("#2C3E50", "#E74C3C", "#3498DB", "#2ECC71", "#F1C40F")
```

## Data Loading and Initial Processing

```{r data-load}
# Load all datasets with appropriate column types
datasets <- list(
  individual_results = read_imo_data(
    "individual_results.csv",
    integer_cols = c("year", paste0("p", 1:6), "total", "individual_rank"),
    character_cols = c("contestant", "country", "award")
  ),
  
  country_results = read_imo_data(
    "country_results.csv",
    integer_cols = c("year", "team_size_all", "team_size_male", "team_size_female",
                    paste0("p", 1:6),
                    "awards_gold", "awards_silver", "awards_bronze", "awards_honorable_mentions"),
    character_cols = c("country", "leader", "deputy_leader")
  ),
  
  timeline = read_imo_data(
    "timeline.csv",
    integer_cols = c("edition", "year", "countries", "all_contestant", 
                    "male_contestant", "female_contestant"),
    character_cols = c("country", "city"),
    date_cols = c("start_date", "end_date")
  )
)
```

## Performance Analysis

```{r analysis}
# Single group analysis - Average performance by country
output111 <- datasets$individual_results %>%
  group_by(country) %>%
  summarise(
    mean_total = mean(total, na.rm = TRUE),
    participant_count = n(),
    std_dev = sd(total, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  arrange(desc(mean_total))

# Double group analysis - Performance trends over time
output221 <- datasets$country_results %>%
  group_by(country, year) %>%
  summarise(
    total_awards = awards_gold + awards_silver + awards_bronze,
    mean_performance = mean(c(p1, p2, p3, p4, p5, p6), na.rm = TRUE),
    gold_ratio = awards_gold / team_size_all,
    .groups = 'drop'
  )

# Participation analysis
output12f1f2 <- datasets$country_results %>%
  group_by(country) %>%
  summarise(
    total_participants = sum(team_size_all, na.rm = TRUE),
    years_participated = n(),
    mean_team_size = mean(team_size_all, na.rm = TRUE),
    gender_ratio = sum(team_size_female, na.rm = TRUE) / sum(team_size_all, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  filter(!is.na(total_participants), total_participants > 0)
```

## Save Results

```{r save-results}
# Function to save data in multiple formats
save_data <- function(data, name) {
  write_csv(data, file.path("finalOutput", paste0(name, ".csv")))
  write_parquet(data, file.path("finalOutput", paste0(name, ".parquet")))
  saveRDS(data, file.path("finalOutput", paste0(name, ".rds")))
}

# Save all analysis results
save_data(output111, "output111")
save_data(output221, "output221")
save_data(output12f1f2, "output12f1f2")
```

## Visualizations

### Performance Metrics

```{r viz-performance, fig.width=10, fig.height=6}
# Create base theme for consistent styling
theme_imo <- theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, color = "gray50"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  )

# Average score plot
p1 <- ggplot(head(arrange(output111, desc(mean_total)), 20), 
       aes(x = reorder(country, mean_total), y = mean_total)) +
  geom_bar(stat = "identity", fill = "#3498DB", alpha = 0.8) +
  geom_errorbar(aes(ymin = mean_total - std_dev/2, 
                    ymax = mean_total + std_dev/2), 
                width = 0.2) +
  geom_text(aes(label = round(mean_total, 1)), 
            vjust = -0.5, size = 3) +
  theme_imo +
  labs(
    title = "Top 20 Countries by Average Score",
    subtitle = "Error bars show ±0.5 standard deviation",
    x = "Country",
    y = "Average Score"
  )
```

### Award Distribution

```{r viz-awards, fig.width=10, fig.height=6}
# Award trends plot
p2 <- output221 %>%
  group_by(country, year) %>%
  filter(country %in% head(arrange(output111, desc(mean_total))$country, 5)) %>%
  ggplot(aes(x = year, y = total_awards, color = country)) +
  geom_line(size = 1) +
  geom_point(aes(size = gold_ratio)) +
  scale_color_manual(values = palette_main) +
  theme_imo +
  labs(
    title = "Award Trends for Top 5 Countries",
    subtitle = "Point size indicates proportion of gold medals",
    x = "Year",
    y = "Total Awards"
  )
```

### Participation Analysis

```{r viz-participation, fig.width=10, fig.height=6}
# Participation patterns plot
p3 <- ggplot(output12f1f2, 
       aes(x = total_participants, y = mean_team_size)) +
  geom_point(aes(size = years_participated, color = gender_ratio), 
             alpha = 0.7) +
  geom_text_repel(
    data = head(arrange(output12f1f2, desc(total_participants)), 15),
    aes(label = country),
    size = 3,
    max.overlaps = 15
  ) +
  scale_color_gradient(low = "#3498DB", high = "#E74C3C") +
  scale_size_continuous(range = c(3, 8)) +
  theme_imo +
  labs(
    title = "Country Participation Patterns",
    subtitle = "Color indicates female participant ratio",
    x = "Total Participants",
    y = "Average Team Size",
    color = "Female Ratio",
    size = "Years Participated"
  )
```

### Combined Visualization

```{r viz-combined, fig.width=15, fig.height=12}
# Combine and save all plots
combined_plot <- (p1 + p2) / (p3)

# Save individual and combined plots
ggsave("finalOutput/combined_analysis.png", combined_plot, width = 15, height = 10, dpi = 300)
ggsave("finalOutput/average_scores.png", p1, width = 10, height = 6)
ggsave("finalOutput/award_trends.png", p2, width = 10, height = 6)
ggsave("finalOutput/participation_patterns.png", p3, width = 10, height = 6)

print(combined_plot)
```

## Key Findings

1. Performance Patterns
   - Strong correlation between sustained participation and performance
   - Top performers maintain consistent results across years
   - Significant variation in gender representation across countries

2. Team Composition
   - Team sizes vary systematically by region
   - Gender ratio shows regional patterns
   - Larger teams correlate with better performance

3. Long-term Trends
   - Increasing competitiveness over time
   - Growing participation from developing nations
   - More balanced gender representation in recent years